Towards a Wide-Coverage Grammar for Swedish\\ Using GF

### 2012-01-13 doc/report/report.tex
report: last comments fixed

We get closer to a satisfying solution by introducing new object type,
\verb_Obj_, which is
identical to normal \verb_NP_ except that it depends on 
the antecedent. All functions where objects are modified in the same way
as other \verb-NP-s are duplicated, one version is dealing 
with objects and one with \verb-NP-.
\begin{verbatim}
ReflVP: Obj     -> VPSlash -> VP ;
     -- sin mat +  äta     -> äta sin mat 
NPtoObj : NP  -> Obj ;
PredVP  : NP -> VP -> Cl ;  -- NP used as subject
\end{verbatim}
Any noun phrase may also be used as an object, whereas subjects only
can be made up by \verb-NP-s.
The drawback of this solution is the duplication of functions.
\begin{verbatim}
PredetNP  : Det -> NP  -> NP ;
PredetObj : Det -> Obj -> Obj ;
\end{verbatim}
%We will need functions for conjugating objects with objects, 
% say smt about overgeneration?
The dependence on the antecedent needs to be applied to adverbial phrases and
adjective phrases as well. The adjective phrase in (\ref{ex:ReflAP}a) and
the adverbial phrase in (\ref{ex:ReflAdv}b) are examples of this; if the
subject was changed to 2nd person, they  would need to change too.
\eenumsentence{%\begin{tabular}[t]{@{}*{5}{l@{\ }}} 
\item\textbf{Du} är längre än \textbf{din} kompis.\\
``You are taller than your friend." 
\item
*\textbf{Du} är här oftare än \textbf{sin} kompis.\\
``You are here more often than {\sc self's} friend."}
\end{tabular}} \label{ex:ReflAP2} \label{ex:ReflAdv2}
Adverbs containing reflexive possessive pronouns may further not be used
to construct subject noun phrases.
%TODO understandable?
\enumsentence{\shortexnt{5}
{*Taket & på & sitt & hus & läcker}
{~~The roof & on & {\sc self's} & house & leaks}}
\vspace{-2mm}
\verb-      (NP    +(Adv (NP Obj)))-\\

@@

The use of dependent types separates our grammar from the other resource grammars. Such separation
is not desirable in itself, but there are at least two reasons why we
believe it is appropriate in this case:
\begin{itemize}
\item The new grammar can be made compatible with the old implementation.
%TODO verify
\item The possibility for noun phrases to agree with an antecedent 
      will be useful in other languages. For example, the Slavic and Romance languages
      have similar reflexive pronouns.
      %and English reflexive pronouns (eg. `himself') cannot
      %be used without antecedent.
\end{itemize}
Moreover, the goal of the project is to make a language specific grammar,
whereas the common abstract is developed to be generalizable enough to 
describe any language. It thus would be surprising if we could cover
all syntactical aspects of one language while leaving the abstract grammar
unchanged.

@@

%{\emph{the opinions}& \emph{that}&\emph{ put+{\textbf s}} & \emph{forward ...}}
%{``The opinions presented ..."}}
%
%A few problematic things can be illuminated:
%Since the constructions are inspired by other languages, some of them may sound quite
%non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
%and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
%\emph{`She sleeps, which is good'} as
%\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
%embedded in the relative clause. This is easily fixed, by adding an extra
%parameter was needed to tell whether \emph{`som'} or \emph{`vilket'} should be used. \\
As a side note, some complications regarding the function \verb-RelCl- can
be pointed out.
%Another complication stems from \verb|RelCl| which express the 
The English implementation of this construction is \emph{`such that'},
and the Swedish version \emph{`sådan att'} sounds awkward, except when used in of
logic and mathematics books.
\eenumsentence{%\begin{tabular}{ll}
\item\emph{From the resource grammar}\\
Jag vill ha en katt sådan att den inte fäller\\
``I want a cat such that it does not shed"
\item
\emph{An alternative formulation} \\ 
Jag vill ha en sådan katt som inte fäller\\
``I want such a cat that it does not shed"}
%\end{tabular}} 
In GF, this relative clause \emph{``en katt sådan att den inte fäller"} is constructed by
two constituents: \emph{`katt'}
of type \verb-CN- and \emph{``den inte fäller"} of type \verb-RCl- (relative clause), glued
together by \emph{``sådan att"}.
As a subject \emph{`den'} is of type \verb-NP- and does not depend on any
agreement. However, it should agree with its antecedent, \emph{`katt'}. 
\enumsentence{*Jag vill ha en katt sådan att \textbf{det} inte fäller.} \label{sent:detfaller}
The dependent types introduced in section \ref{sec:reflexives} could probably be
useful.
However a carefully prepared analysis and motivation is needed, %ationamount in a bigger reconstruction which therefore
%would require a. 
and sentence (\ref{sent:detfaller}) is still accepted by the grammar.

@@

\subsubsection{An example of overgeneration and how it can be solved}
The \verb-Extra- module contains a function for constructing genitives.
\begin{verbatim}GenNP : NP -> Quant ;\end{verbatim}
It selects the possessive form given in the \verb-NP- and changed its type
to \verb-Quantifier-. 
%\begin{verbatim}
%(DetCN (DetQuant (GenNP (DetCN (DetQuant DefArt NumSg) (UseN boy_N))) NumSg) (UseN cat_N))
%\end{verbatim}
The \verb-GenNP- function %was %severely
overgenerated when this project started, even
though its implementation was not erroneous in itself. The cause of the overgeneration 
was a combination of two things:
\begin{itemize}
\item \verb-Quantifiers- do not have any possessive
forms, so when a quantifier is turned into
a \verb-NP- (by the functions \verb-DetQuant- and \verb-DetNP-), its genitive field gets
the same form as its nominative field.
\item\verb-GenNP-
introduced a cycle: a \verb-Quant- may be turned into a \verb-NP-, and now also
a \verb-NP- can be turned into a \verb-Quant-.\\
\verb|GenNP : NP -> Quant, DetQuant : Quant -> Det; DetNP : Det -> NP|
\end{itemize}
We had that a noun phrase, eg \emph{`katten'} could be turned into a quantifier,
by using its genitive form, \emph{``kattens'}. This could be used as a determiner,
\newpage % until behave
\enumsentence{Det är kattens mat.\\``It is the cat's food."}
and thus as a noun phrase:
\enumsentence{Det är kattens.\\``It is the cat's."}
The possessive form of this \verb-NP- is now the same as its nominative:
\emph{`kattens'}. Hence, the function \verb-GenNP- could be applied
any number of times without any change in the output.

@@

\subsubsection{Using Talbanken to assign correct categorisations}
\label{sec:gf.quant}
One of the biggest differences between the GF analysis and the one made in Talbanken
is related to pronouns. 
As noted in section \ref{sec:saldoRes} our GF grammar only classifies personal pronouns
as pronouns. Many pronouns could not be imported from Saldo, since the correct category
could not be identified automatically. Therefore, some experiments were
conducted on how to extract this 
information from Talbanken. We were interested in words which can be used
the same ways as `sådana' (\emph{`such'}) that acts either as a
determiner, as in sentence (\ref{ex:QuantDet}), or as an adjective as
 in sentence (\ref{ex:QuantAdj}). 
\eenumsentence{
\item\shortex{5}
{Sådana& \label{ex:QuantDet}katter& vill& jag& ha}
{\emph{such}& \label{ex:QuantDet}\emph{cats}& \emph{want}& \emph{I}& \emph{have}}
{``I want such cats'}
\item\shortex{3}
{Han& är& \label{ex:QuantAdj}sådan\toplabel{sent:Quant}}
{\emph{he}& \emph{is}& \label{ex:QuantAdj}\emph{such}\toplabel{sent:Quant}}
{``He's like that"}
}

@@

\chapter{Extracting a GF treebank} %bootstraping/extracting a GF treebank
\label{sec:Mapping}
%>> no subjects or object, analyse is more from the lexicon,
%>> compare to others, ask elisabet.
%>> diffs: valencies, pp (gf ingen skillnad direkt/indirekt i träd, pga multiling)
%>> mamba very used and tested for swedish, standard.
%>> focus on what we DO cover.\\
Talbanken contains much valuable information about phrase structures and word 
usage. It is analyzed with the MAMBA annotation scheme (Telemann, 1974).% which is 
%>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
%>> for large scale projects.
%annotation scheme
%commonly used, and hence very well-tested, for describing Swedish.
One part of this project has focused on translating trees from Talbanken to GF
by constructing a mapping, which 
automatically transforms trees in the Talbanken format to GF abstract syntax trees.
We hence get a comparison between the two annotations and at the same time
we get methods for extracting and translating to GF notation. 
Figure \ref{fig:translationtrees} shows an example of a visualized Talbanken05 tree
of the sentence ``Katten på bilen blir större" (\emph{``The cat on the car gets bigger"})
and the result of translating it to GF.
Both the POS tags as well as the syntactic information are needed
during the translation, so that all the information shown in the abstract tree
(\ref{pic:gftree}) can be extracted. We need to know that \emph{katten} is a noun
used in definite form (\verb-DefArt-), singular (\verb-NumSg-). Even though the
notation is translated, the original analyse from Talbanken
is preserved.
\emph{på bilen} should still form a subtree attached to \emph{katten}.
%produces a \verb-NP- from `katten' this means that
%DetCN : Quant                  -> CN    -> NP
%DetCN  (DetQuant DefArt NumSg)    cat_N = katten

@@

%The mapping works for Talbanken05 in Tiger XML-format.
\section{The Talbanken annotation}
Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give part of sentence information,
such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the grammatical functions: \verb|SS| for subject, 
\verb|OO| for object etc. Nearly 400 different tags are used, of which more than
300 different are POS-tags. The high number is due to the level of detail,  there are for
example almost 50 tags for nouns, excluding proper names. The tags
show definiteness, case and whether the word is a compound. Some words also have
their own tags.\\

@@

\section{The translation process}
The implementation consists of a set of rules that translates
one node of a Talbanken tree to a GF function, possibly by describing what the
nodes subtree should look like. 
\begin{verbatim}
translate S  = do
             np <- find NP
             vp <- find VP
             return (PredVP np vp)
\end{verbatim}
This rule tells that we may map the category \verb-S- to the function 
\verb-PredVP-, given that we find a noun phrase and a verb phrase in
the subtree of \verb-S-.

@@

Lets look at the example in figure
\ref{pic:tbtree}. The verb \emph{`bli'} is
followed by a predicative 
adjective modifying the subject, \verb|SP|. GF admits \emph{`bli'} to have
type \verb|VA|, a verb with an adjective as complement, and the translator can
apply the GF function \verb|ComplVA| to combine the verb with its object. \\

@@

\noindent since the distinction between different types of verbs is not always clear in Swedish;
most transitive verbs can also be used as intransitive. Both sentences in
example \ref{ex:mappTransitive} are grammatically correct but only (\ref{ex:mappTransitive}b)
is accepted by the GF grammar, and
the grammar rules are simply too strict for accepting many Talbanken sentences.
\enumsentence{\begin{tabular}[t]{@{}*{14}{l@{\ }}}
a. &Han& sitter& och &läser.&& & b. &Han& sitter& och& läser& en& bok.\\
&\emph{he}& \emph{sits}& \emph{and} &\emph{reads} &&&
&\emph{he}& \emph{sits}& \emph{and}& \emph{reads}& \emph{a}& \emph{book}\\
\end{tabular}}
\label{ex:mappTransitive}
See section \ref{sec:futureValency} for a longer discussion about this.\\
%the Talbankenwill allow words to be used in other ways than their
%GF valencies prescribes.

@@

Some more information about VP are given by the tag \verb|VG| (\emph{verb group}),
which groups an infinitival verb and its object together into a \verb|VP|.
This information can however be extracted from the flat version, and the results
get slightly better when using the flat version. \\

When evaluating the mapping, the results strongly depend on which restrictions we
put on the input. 
One of the reasons why a node cannot be translated, is the 
use of the tags show in figure \ref{fig:mapBadtag}.
The \textbf{PU} tag is used for graphic listings, and not for fluent text.
In our grammar there is naturally no corresponding function; 
%We are not interested in sentences made up by list, since 
the listings are meant for making the text look nice in
folders etc and outside the scope for the grammar itself. The tags \textbf{XX} and \textbf{NP} are often used since
Talbanken -- unlike GF -- makes a difference between subjects
and objects. %(which is not needed in GF (see section \ref{sec:swegf}))
The analysis of elliptical expression in (\ref{sent:krav})
\enumsentence{För stora krav.\\
``Too high demands."}\label{sent:krav}
contains the tags \textbf{XX} and \textbf{NP}, since it is not obvious
whether the noun phrase is used as subject or an object.
The tags shown in figure \ref{fig:mapBadtag} occur quite frequently in the treebank and are always translated
to metas, which lowers our result. \\
%How we cannot cover all cases. 
\begin{figure}[h]
\begin{tabular}{ll}
\textbf{NAC} & Not a constituent\\
\textbf{XP} & Other (non-coordinated) phrase\\
\textbf{DB} & Doubled function\\
\textbf{PU} & List item\\
\textbf{XX} & Unclassifiable part-of-speech\\
\end{tabular}
\caption{}\label{fig:mapBadtag}
\end{figure}

@@

A mapping between GF and the Wall Street Journal of Penn Treebank has earlier been conducted \cite{gfMech}.
The percentage of restored nodes from Peen Treebank is higher than our results.
%TODO review
The reason for this may be the fact that English is syntactically less complicated
than Swedish.
Furthermore, the text in Talbanken are from various brochures, newspapers and text books,
where idiomatic expressions are more likely and the language presumably less strict than
in Wall Street Journal.
Also, the Penn Treebank contains a lower number of tags, 82 compared
to more than 300 in Talbanken. Even if the tags describing the same word
%TODO review
class as another tag are excluded, Talbanken still leaves us with more than 130 tags.
With more tags, we get more information, but as the number 
%Thus the number of
increase, so does the amount of work of finding the correct translation 
for each combination of tags and writing rules that cover all constructions. 
%iand so is the number of
%rules that would be needed to cover all constructions.

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which is fast even with an extensive lexicon.
%TODO make nice!
Although it is not been verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantic knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
%TODO move to Misc?
the linearization should be correct. There are some cases when the correctness
of the output has been put aside in order to increase the expressivity, such
as \verb-ComplBareVV-, which use a verb-complement verb without the infinitive mark, 
as in sentence (\ref{sent:complVV1}b).
\eenumsentence{\item\shortex{5}
{Jag& börjar&\textbf{att}& bli& hungrig}
{\emph{I}& \emph{begin}& \textbf{\emph{to}}&\emph{become}& \emph{hungry}}
{``I'm getting hungry"}
\item
\shortex{4}
{Jag& börjar& bli& hungrig}
{\emph{I}& \emph{begin}& \emph{become}& \emph{hungry}}
{``I'm getting hungry"}}\label{sent:complVV1}
Since there is no information about which verbs that allows the infinitive mark
to be left out, it will also allow the more questionable sentence (\ref{sent:complVV2}b).
\eenumsentence{
\item \shortex{5}
{Jag & låter &bli & \textbf{att}& titta} 
{\emph{I} & \emph{let} & \emph{be} &\textbf{\emph{to}} &\emph{watch}}
{``I refrain from watching"}
\item \shortexnt{4}
{*Jag & låter & bli & titta} 
{~~\emph{I} & \emph{leave} &\emph{be} & \emph{watch}}}\label{sent:complVV2}
As these functions provide stylistic flexiblity when parsing, they can be left out
when generating language, and the grammar then provides text of good
grammatical quality. \\
%ComplBareVV, SupCl
%For random generation, the sentences generated may be next to impossible
%to interpret without looking at the abstract tree, since they may contain
%too many attributes, relative clauses etc.

@@

Parsing bigger parts of Talbanken would hence give very low results at this stage, 
%, for the reasons stated.
and a comparison of the results would not be of much value as
there would not be enough material %underlag is yet too small 
to do be able to do any interesting analysis.
An estimation of the improvement can be given by looking at the results from
running the test suite used for the grammar development.
The sentences given in the test suite are short, four to 10 words, and the
words are in most cases included in
the lexicon, but there are also constructions that have not been implemented 
during this project.
The first grammar could parse almost half of the sentences, the result for the final
grammar was 66 \%. 
% try NPs?
It is thus not yet interesting to talk about coverage, but about the quality and
the ability to scale up, which has so far proved to be good.
I further believe that the presence of a expert in Swedish, professor Elisabet
Engdahl\footnote{http://svenska.gu.se/om-oss/personal/elisabet-engdahl},
has
increased the standard substantially.\\

@@

\subsection{Enhancements and complements}
\subsubsection{Grammar}
The grammar should cover the most prominent Swedish features.
%All imaginable grammatical constructs cannot be cover by our grammar.
Some work must be left for the robustness, but there are specific constructs
that are desirable and interesting to implement. A few examples are listed here:\\

@@

\subsection{Making the parser robust}
To achieve our goal of parsing unrestricted text, we need to use statistical methods.
The parser needs to be able to handle
ellipses and long sentences, names and numbers,
compound words and unknown constructions.

>>>>

We get closer to a satisfying solution by introducing new object type,
\verb_Obj_, which is
identical to normal \verb_NP_ except that it depends on 
the antecedent. All functions where objects are modified in the same way
as other \verb-NP-s are duplicated, one version is dealing 
with objects and one with \verb-NP-.
\begin{verbatim}
ReflVP: Obj     -> VPSlash -> VP ;
     -- sin mat +  äta     -> äta sin mat 
NPtoObj : NP  -> Obj ;
PredVP  : NP -> VP -> Cl ;  -- NP used as subject
\end{verbatim}
Any noun phrase may also be used as an object, whereas subjects only
can be made up by \verb-NP-s.
The drawback of this solution is the duplication of functions.
\begin{verbatim}
PredetNP  : Det -> NP  -> NP ;
PredetObj : Det -> Obj -> Obj ;
\end{verbatim}
%We will need functions for conjugating objects with objects, 
% say smt about overgeneration?
The dependence on the antecedent needs to be applied to adverbial phrases and
adjective phrases as well. The adjective phrase in (\ref{ex:ReflAP}a) and
the adverbial phrase in (\ref{ex:ReflAdv}b) are examples of this; if the
subject was changed to 2nd person, they  would need to change too.
\eenumsentence{%\begin{tabular}[t]{@{}*{5}{l@{\ }}} 
\item\textbf{Du} är längre än \textbf{din} kompis.\\
``You are taller than your friend." 
\item
*\textbf{Du} är här oftare än \textbf{sin} kompis.\\
``You are here more often than {\sc self's} friend."}
 \label{ex:ReflAP2} \label{ex:ReflAdv2}
Adverbs containing reflexive possessive pronouns may further not be used
to construct subject noun phrases.
%TODO understandable?
\enumsentence{\shortexnt{5}
{*Taket & på & sitt & hus & läcker}
{~~The roof & on & {\sc self's} & house & leaks}}
\vspace{-2mm}
\verb-      (NP    +(Adv (NP Obj)))-\\

@@

The use of dependent types separates our grammar from the other resource grammars. Such separation
is not desirable in itself, but there are at least two reasons why we
believe it is appropriate in this case:
\begin{itemize}
\item The new grammar can be made compatible with the old implementation.
%TODO verify
\item The possibility for noun phrases to agree with an antecedent 
      will be useful in other languages. For example, the Slavic and Romance languages
      have similar reflexive pronouns.
      %and English reflexive pronouns (eg. `himself') cannot
      %be used without antecedent.
\end{itemize}
Moreover, the goal of the project is to make a language specific grammar,
whereas the common abstract is developed to be generalizable enough to 
describe any language. It thus would be surprising if we could cover
all syntactical aspects of Swedish while leaving the abstract grammar
unchanged.

@@

%{\emph{the opinions}& \emph{that}&\emph{ put+{\textbf s}} & \emph{forward ...}}
%{``The opinions presented ..."}}
%
%A few problematic things can be illuminated:
%Since the constructions are inspired by other languages, some of them may sound quite
%non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
%and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
%\emph{`She sleeps, which is good'} as
%\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
%embedded in the relative clause. This is easily fixed, by adding an extra
%parameter was needed to tell whether \emph{`som'} or \emph{`vilket'} should be used. \\
As a side note, some complications regarding the function \verb-RelCl- can
be pointed out.
%Another complication stems from \verb|RelCl| which express the 
The English implementation of this construction is \emph{`such that'},
and the Swedish version \emph{`sådan att'} sounds awkward, except when used in of
logic and mathematics books.
\eenumsentence{%\begin{tabular}{ll}
\item\emph{From the resource grammar}\\
Jag vill ha en katt sådan att den inte fäller\\
``I want a cat such that it does not shed"
\item
\emph{An alternative formulation} \\ 
Jag vill ha en sådan katt som inte fäller\\
``I want such a cat that it does not shed"}
%\end{tabular}} 
In GF, this relative clause \emph{``en katt sådan att den inte fäller"} is constructed by
two constituents: \emph{`katt'}
of type \verb-CN- and \emph{``den inte fäller"} of type \verb-RCl- (relative clause), glued
together by \emph{``sådan att"}.
As a subject \emph{`den'} is of type \verb-NP- and does not depend on any
agreement. However, it should agree with its antecedent, \emph{`katt'}. 
\enumsentence{*Jag vill ha en katt sådan att \textbf{det} inte fäller.} \label{sent:detfaller}
The dependent types introduced in section \ref{sec:reflexives} could probably be
useful to solve this problem.
However a carefully prepared analysis and motivation is needed, %ationamount in a bigger reconstruction which therefore
%would require a. 
and sentence (\ref{sent:detfaller}) is still accepted by the grammar.

@@

\subsubsection{An example of overgeneration and how it can be solved}
The \verb-Extra- module contains a function for constructing genitives.
\begin{verbatim}GenNP : NP -> Quant ;\end{verbatim}
It selects the possessive form given in the \verb-NP- and changes its type
to \verb-Quantifier-. 
%\begin{verbatim}
%(DetCN (DetQuant (GenNP (DetCN (DetQuant DefArt NumSg) (UseN boy_N))) NumSg) (UseN cat_N))
%\end{verbatim}
The \verb-GenNP- function %was %severely
overgenerated when this project started, even
though its implementation was not erroneous in itself. The cause of the overgeneration 
was a combination of two things:
\begin{itemize}
\item \verb-Quantifiers- do not have any possessive
forms, so when a quantifier is turned into
a \verb-NP- (by the functions \verb-DetQuant- and \verb-DetNP-), its genitive field gets
the same form as its nominative field.
\item\verb-GenNP-
introduced a cycle: a \verb-Quant- may be turned into a \verb-NP-, and now
a \verb-NP- can be turned into a \verb-Quant-.\\
\verb|GenNP : NP -> Quant, DetQuant : Quant -> Det; DetNP : Det -> NP|
\end{itemize}
We had that a noun phrase, eg \emph{`katten'} could be turned into a quantifier,
by using its genitive form, \emph{``kattens'}. This could be used as a determiner,
\newpage % until behave
\enumsentence{Det är kattens mat.\\``It is the cat's food."}
and thus as a noun phrase:
\enumsentence{Det är kattens.\\``It is the cat's."}
The possessive form of this \verb-NP- is now the same as its nominative:
\emph{`kattens'}. Hence, the function \verb-GenNP- could be applied
any number of times without any change in the output.

@@

\subsubsection{Using Talbanken to assign correct categorisations}
\label{sec:gf.quant}
One of the biggest differences between the GF analysis and the one made in Talbanken
is related to pronouns. 
As noted in section \ref{sec:gfnouns} our GF grammar only classifies personal pronouns
as pronouns. Many pronouns could not be imported from Saldo, since the correct category
could not be identified automatically. Therefore, some experiments were
conducted on how to extract this 
information from Talbanken. We were interested in words which can be used
the same ways as `sådana' (\emph{`such'}) that acts either as a
determiner (\ref{ex:QuantDet}), or as an adjective agreeing with the noun
(\ref{ex:QuantAdj}-c). 
\eenumsentence{
\item\shortex{5}
{Sådana& \label{ex:QuantDet}katter& vill& jag& ha}
{\emph{such}& \label{ex:QuantDet}\emph{cats}& \emph{want}& \emph{I}& \emph{have}}
{``I want such cats'}
\item\shortex{3}
{Han& är& \label{ex:QuantAdj}sådan\toplabel{sent:Quant}}
{\emph{he}& \emph{is}& \label{ex:QuantAdj}\emph{such}\toplabel{sent:Quant}}
{``He's like that"}
\item\shortex{3}
{Huset& är& \label{ex:QuantAdj2}sådan\textbf{t}}
{\emph{the house}& \emph{is}& \label{ex:QuantAdj2}\emph{such}}
{``The house is like that"}
}

@@

\chapter{Extracting a GF treebank} %bootstraping/extracting a GF treebank
\label{sec:Mapping}
%>> no subjects or object, analyse is more from the lexicon,
%>> compare to others, ask elisabet.
%>> diffs: valencies, pp (gf ingen skillnad direkt/indirekt i träd, pga multiling)
%>> mamba very used and tested for swedish, standard.
%>> focus on what we DO cover.\\
Talbanken contains much valuable information about phrase structures and word 
usage. It is analyzed with the MAMBA annotation scheme (Telemann, 1974).
% which is 
%>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
%>> for large scale projects.
%annotation scheme
%commonly used, and hence very well-tested, for describing Swedish.
One part of this project has focused on translating trees from Talbanken to GF
by constructing a mapping, which 
automatically transforms trees in the Talbanken format to GF abstract syntax trees.
We hence get a comparison between the two annotations and at the same time
we get methods for extracting and translating to GF notation. 
Figure \ref{fig:translationtrees} shows an example of a visualized Talbanken05 tree
of the sentence ``Katten på bilen blir större" (\emph{``The cat on the car gets bigger"})
and the result of translating it to GF.
Both the POS tags as well as the syntactic information are needed
during the translation, so that all the information shown in the abstract tree
(\ref{pic:gftree}) can be extracted. We need to know that \emph{katten} is a noun
used in definite form (\verb-DefArt-), singular (\verb-NumSg-). Even though the
notation is translated, the original analyse from Talbanken
is preserved.
\emph{på bilen} should still form a subtree attached to \emph{katten}.
%produces a \verb-NP- from `katten' this means that
%DetCN : Quant                  -> CN    -> NP
%DetCN  (DetQuant DefArt NumSg)    cat_N = katten

@@

%The mapping works for Talbanken05 in Tiger XML-format.
\section{The Talbanken annotation}
Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatictal information,
such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the
part of sentence: \verb|SS| for subject, 
\verb|OO| for object etc. Nearly 400 different tags are used, of which more than
300 different are POS-tags. The high number is due to the level of detail,  there are for
example almost 50 tags for nouns, excluding proper names. The tags
show definiteness, case and whether the word is a compound. Some words also have
their own tags.\\

@@

\section{The translation process}
The implementation consists of a set of rules that translate
nodes of a Talbanken tree to GF functions, possibly by describing what the
nodes subtree should look like. 
\begin{verbatim}
translate S  = do
             np <- find NP
             vp <- find VP
             return (PredVP np vp)
\end{verbatim}
This rule tells that we may map the category \verb-S- to the function 
\verb-PredVP-, given that we find a noun phrase and a verb phrase in
the subtree of \verb-S-.

@@

%Lets look at the example 
In figure \ref{pic:tbtree} the verb \emph{`bli'} is
followed by a predicative 
adjective modifying the subject tagged with \verb|SP|. Our GF grammar admits
\emph{`bli'} to have type \verb|VA|, a verb with an adjective as complement,
and the translator can hence apply the GF function \verb|ComplVA| to combine the verb
with its object. \\

@@

\noindent since the distinction between different types of verbs is not always clear in Swedish;
most transitive verbs can also be used as intransitive. Both sentences in
example (\ref{ex:mappTransitive}) are grammatically correct but only (\ref{ex:mappTransitive}b)
is accepted by the GF grammar: 
the grammar rules are simply too strict for accepting many Talbanken sentences.
\enumsentence{\begin{tabular}[t]{@{}*{14}{l@{\ }}}
a. &Han& sitter& och &läser.&& & b. &Han& sitter& och& läser& en& bok.\\
&\emph{he}& \emph{sits}& \emph{and} &\emph{reads} &&&
&\emph{he}& \emph{sits}& \emph{and}& \emph{reads}& \emph{a}& \emph{book}\\
\end{tabular}}
\label{ex:mappTransitive}
%See section \ref{sec:futureValency} for a longer discussion about this.\\
%the Talbankenwill allow words to be used in other ways than their
%GF valencies prescribes.

@@

Some more information about VP are given by the tag \verb|VG| (\emph{verb group}),
which groups an infinitival verb and its object together into a \verb|VP|.
This information can however be extracted from the flat implementation, and the results
get slightly better when using this version. \\

When evaluating the mapping, the results strongly depend on which restrictions we
put on the input. 
One of the reasons why a node cannot be translated, is the 
use of the tags show in figure \ref{fig:mapBadtag}.
The \textbf{PU} tag is used for graphic listings, and not for fluent text.
In our grammar there is naturally no corresponding function; 
%We are not interested in sentences made up by list, since 
the listings are meant for making the text look nice in
folders etc and outside the scope for the grammar itself. The tags \textbf{XX}
and \textbf{NAC} are often used since Talbanken makes a
difference between subject
and object noun phrases. %(which is not needed in GF (see section \ref{sec:swegf}))
The analysis of elliptical expression in (\ref{sent:krav})
\enumsentence{För stora krav.\\
``Too high demands."}\label{sent:krav}
contains the tags \textbf{XX} and \textbf{NAC}, since it is not obvious
whether the noun phrase is used as subject or an object.
The tags shown in figure \ref{fig:mapBadtag} occur quite frequently in the treebank and are always translated
to metas, which lowers our result. \\
%How we cannot cover all cases. 
\begin{figure}[h]
\begin{tabular}{ll}
\textbf{NAC} & Not a constituent\\
\textbf{XP} & Other (non-coordinated) phrase\\
\textbf{DB} & Doubled function\\
\textbf{PU} & List item\\
\textbf{XX} & Unclassifiable part-of-speech\\
\end{tabular}
\caption{}\label{fig:mapBadtag}
\end{figure}

@@

A mapping between GF and the Wall Street Journal of Penn Treebank has earlier been conducted \cite{gfMech}.
The percentage of restored nodes from Peen Treebank is higher than our results.
%TODO review
The reason for this may be the fact that English is syntactically less complicated
than Swedish.
Furthermore, the text in Talbanken are from various brochures, newspapers and text books,
where idiomatic expressions are more likely the be appear and the language
presumably less strict than in Wall Street Journal\footnote{www.wsj.com}.
Also, the Penn Treebank contains a lower number of tags, 82 compared
to more than 300 in Talbanken. Even if the tags describing the same word
%TODO review
class as another tag are excluded, Talbanken still leaves us with more than 130 tags.
With more tags, we get more information, but as the number 
%Thus the number of
increase, so does the amount of work of finding the correct translation 
for each combination of tags and writing rules that cover all constructions. 
%iand so is the number of
%rules that would be needed to cover all constructions.

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which works fast even when using an extensive lexicon.
%TODO make nice!
Although it is not been formally verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantic knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
%TODO move to Misc?
the linearization should be correct. There are some cases when the correctness
of the output has been put aside in order to increase the expressivity, such
as \verb-ComplBareVV-, which use a verb-complement verb without the infinitive mark, 
as in sentence (\ref{sent:complVV1}b).
\eenumsentence{\item\shortex{5}
{Jag& börjar&\textbf{att}& bli& hungrig}
{\emph{I}& \emph{begin}& \textbf{\emph{to}}&\emph{become}& \emph{hungry}}
{``I'm getting hungry"}
\item
\shortex{4}
{Jag& börjar& bli& hungrig}
{\emph{I}& \emph{begin}& \emph{become}& \emph{hungry}}
{``I'm getting hungry"}}\label{sent:complVV1}
Since there is no information in the grammar or lexicon about which verbs that
allows the infinitive mark to be left out, it will also allow the more
questionable sentence (\ref{sent:complVV2}b).
\eenumsentence{
\item \shortex{5}
{Jag & låter &bli & \textbf{att}& titta} 
{\emph{I} & \emph{let} & \emph{be} &\textbf{\emph{to}} &\emph{watch}}
{``I refrain from watching"}
\item \shortexnt{4}
{*Jag & låter & bli & titta} 
{~~\emph{I} & \emph{leave} &\emph{be} & \emph{watch}}}\label{sent:complVV2}
As these functions serve to provide stylistic flexiblity when parsing, they can be left out
when generating language, and the grammar then generates text of good
grammatical quality. \\
%ComplBareVV, SupCl
%For random generation, the sentences generated may be next to impossible
%to interpret without looking at the abstract tree, since they may contain
%too many attributes, relative clauses etc.

@@

Parsing bigger parts of Talbanken would hence give very low results at this stage, 
%, for the reasons stated.
and a comparison of the results would not be of much value as
there would not be enough material %underlag is yet too small 
to do be able to do any interesting analysis.
An estimation of the improvement can be given by looking at the results from
running the test suite used for the grammar development.
The sentences given in the test suite are short, four to 10 words, and the
words are in most cases included in
the lexicon, but there are also constructions that have not been implemented 
during this project.
The first grammar could parse less than half of the sentences, the result for the final
grammar was 66 \%. 
% try NPs?
It is thus not yet interesting to talk about coverage, but about the quality and
the ability to scale up, which has so far proved to be good.
I further believe that the presence of a expert in Swedish, professor Elisabet
Engdahl\footnote{http://svenska.gu.se/om-oss/personal/elisabet-engdahl},
has
increased the standard substantially.\\

@@

\subsection{Enhancements and complements}
\subsubsection{Grammar}
The grammar should cover the most prominent Swedish features.
%All imaginable grammatical constructs cannot be cover by our grammar.
Even as some work must be left for the robustness, there are specific constructs
that are desirable and interesting to implement. A few examples are listed here:\\

@@

\subsection{Making the parser robust}
To achieve our goal of parsing unrestricted text, we need to use statistical methods.
The parser needs to be able to handle
ellipses and long and complex sentences, names and numbers,
compound words and other unknown constructions.

\subsubsection{Parallel parsing/Chunk parsing}
Instead of parsing a whole sentence in one go, we intend
parse parts of it in parallel and then try to combine the pieces, building
the sentence bottom up.
This method should be faster then normal parsing and would
at the same time give robustness;
components that cannot be combined by any grammar rule are combined by a 
meta function. When an unknown construction is encountered, the parser
will not fail but give a partial parse tree with the analysis for the rest of
the sentence \cite{gfMech}.
The output given to the user will be more valuable, since partial parse
trees are more interesting the no parse trees.
%We might end up with partial parse trees, but we would
%be able to parse at least parts of every sentence and give a more
%interesting result to the user.
